Exploring TTC Delays

### 2024-01-25 outputs/paper/paper.qmd
Cleaned data and started writing paper.

---
title: "Exploring TTC Delays"
subtitle: "A Multimodal Analysis"
author: Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé coffee, Rajan Maghera, Egg McMuffins, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you. No thanks go to Rohan in particular for not pointing out all the R he wanted us to learn that was in his textbook... sigh. I will try the appendix cocktails someday, though. :)"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
fig_caption: yes
number-sections: true
bibliography: references.bib
---

@@

library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(data.table)
library(dplyr)
library(kableExtra)
```

@@

A key area of improvement in Canadian sustainability is in increasing the use of public transportation in big cities like Toronto. This makes sense since cars account for about 80% of transportation emissions [@diaetal]. However, a major drawback preventing the widespread use of public transportation is the existence of delays within the transit system. These delays incentivize citizens toward using less sustainable alternatives like private cars, leading to congestion on highways and consequent increased emissions. The lack of timely transit services have a demonstrated impact on traffic. One study by Anderson et al. estimated a whopping 47 percent increase in highway delay when transit service ceased completely in a big city such as Los Angeles [@anderson]. It is clear to see, then, that it is imperative that policy makers understand and fix these delays to improve sustainability.

The "TTC", or Toronto Transit Commission, has been Toronto's transit system since 1921, serving the city population with subways, streetcars, buses, along with smaller offerings. The openly available data surrounding the delays of these various modes of transport offers opportunity to investigate the root causes of the longer-lasting delays. Understanding what caused a delay, when it was caused and where it was caused could give the TTC more insight into how to deliver their services to the public more efficiently and effectively, setting an examples of what an effective transit agency could look like.

This paper aims to find relationships and patterns of public transit delay time across various genres of transportation. First, the data source and analysis employed are covered. Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration.

@@

To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 was downloaded using the OpenDataToronto R package [@odt]. Then, the data was cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages. After combining the datasets together, variables common between datasets and relevant to the analysis were selected. In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened. (Perhaps add seasonal analysis??)

## General Delay Analysis {#subsec1-data}

@@

sample_table <- head(cleaned_transit_data, 5)

knitr::kable(sample_table)
```

>>>>

---
title: "Exploring TTC Delays"
subtitle: "A Multimodal Analysis"
author: Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you. No thanks go to Rohan in particular for not pointing out all the R he wanted us to learn that was in his textbook... sigh. I will try the appendix cocktails someday, though. :)"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
fig_caption: yes
number-sections: true
bibliography: references.bib
---

@@

library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(data.table)
library(dplyr)
library(kableExtra)
library(knitr)
```

@@

A key area of improvement in Canadian sustainability is in increasing the use of public transportation in big cities like Toronto. This makes sense, since cars account for about 80% of transportation emissions [@diaetal]. A major drawback preventing the widespread use of public transportation is the existence of delays within transit systems. These delays push citizens toward using less sustainable alternatives like private cars, leading to congestion on highways and consequent increased emissions. The lack of timely transit services have a demonstrated impact on traffic. One study by Anderson et al. estimated a whopping 47 percent increase in highway delay when transit service ceased completely in a big city such as Los Angeles [@anderson]. Learning from this example, it is imperative that policy makers understand and fix these delays to improve sustainability in Toronto.

The Toronto Transit Commission (TTC) has been Toronto's transit system since 1921, serving the city population with a variety of transportation methods, particularly subways, streetcars, and buses. The openly available data surrounding the delays of these various modes of transport offers the opportunity to investigate the root causes of these delays and prompt discussion as how to rectify them. Understanding the common causes of a delay, when a delay is typically caused and where it was caused gives the TTC more insight into how to deliver their services to the public more efficiently and effectively, setting an improved example of what an effective transit agency looks like.

This paper aims to find patterns of public transit delay time across various genres of transportation. First, the data source and analysis employed are covered in Section Two. Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration for this data.

@@

To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 was downloaded using the OpenDataToronto R package [@odt]. Then, the data was cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages. After combining the datasets together, variables common between datasets and relevant to the analysis were selected, leading to a dataset with 93,569 observations. In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened.

@@

```

```{r}
#| message: false
#| echo: false
#| label: summary_stats_table
#| tbl-cap: Sample of Cleaned Delay Data
#| fig-align: center

sample_table <- cleaned_transit_data |>
  rename(Vehicle = vehicle) |>
  rename(`Delay (Minutes)` = `Min Delay`) |>
  rename(Reason = Incident) |>
  rename(`Route/Line` = Line) |>
  arrange(Date)

sample_table <- head(sample_table, 5)

#Scaling taken from: 
#https://stackoverflow.com/questions/56254631/table-way-too-wide-to-fit-in-markdown-generated-pdf
kable(sample_table) |> kable_styling(latex_options = c("striped", "scale_down"))
```

```{r}
#| message: false
#| echo: false
#| label: common-reasons-delay
#| tbl-cap: Most Common Reasons for Delay on TTC

#Bar chart of Reasons for delay vs. # of incidents

cleaned_transit_data |>
  count(Incident) |>
  ggplot(aes(x = n, y = reorder(Incident, n))) +
  geom_bar(stat="identity", fill="grey50") +
  labs(title="Most Common Reasons for Delay", x="Count", y="Incident Type") +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5),
        panel.background = element_blank(),
        axis.line.y=element_line(color="black"),
        axis.ticks.y=element_blank(),
        panel.grid.major.x=element_line(color="grey80"),
        panel.grid.minor=element_blank())

```

```{r}
#| message: false
#| echo: false
#| label: monthly-delay-times
#| tbl-cap: Average Monthy Delay Times
#Line chart of how avg. delay time changes over the months

month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)

# #Calculate avg. delays 
avg_delay <- month_dataset |> group_by(month, vehicle) |> summarise(`Min Delay` = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")

# #convert month to factor (categorical data)
avg_delay$month <- factor(avg_delay$month, levels = 1:12, labels = month.abb[1:12])

# #plot the graph
avg_delay |>
  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
  geom_line() +
  labs(title="Average Monthly Delays by Vehicle", x="Month", y="Monthly Avg. Delay") +
  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
  theme_minimal()
```

```{r}
#| message: false
#| echo: false
#| label: mode-longest-delays
#| tbl-cap: Most Common Reasons for Delay on TTC
#Bar graph of mode of transport with longest delays

avg_time_delayed_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal()
```

```{r}
#| message: false
#| echo: false
#| label: most-incidents-vehicle
#| tbl-cap: Most Incidents by Vehicle
#Bar graph of mode of transport with most number of delays

num_incidents_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(num_incidents = n(), .groups = "drop")

num_incidents_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = num_incidents)) +
  geom_col() + 
  labs(title="Number of Incidents by Vehicle", x="Vehicle Type", y="Number of Incidents") +
  theme_minimal()

```

```{r}
#| message: false
#| echo: false
#| label: transport-delay-time
#| tbl-cap: Most Common Reasons for Delay on TTC
#Bar graph of mode of transport with most number of delays

# Inspired by: https://stackoverflow.com/questions/34280140/order-bar-graph-in-order-in-this-case-as-per-days-of-week

clean_transit_data <- cleaned_transit_data

# Define the order of days
days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

# Convert 'Day' column to a factor with specific levels
clean_transit_data$Day <- factor(clean_transit_data$Day, levels = days_order)

# Now create the plot
num_incidents_vehicle <- clean_transit_data |> arrange(Day) |> group_by(Day) |> summarise(num_incidents = n(), .groups = "drop")

num_incidents_vehicle |>
  ggplot(mapping = aes(x = Day, y = num_incidents)) +
  geom_col() +
  labs(title="Count of Delays by Day of Week", x="Day of Week", y="Incident Count") +
  theme_minimal()

```

```{r}
#| message: false
#| echo: false
#| label: transport-delay-number
#| tbl-cap: Most Common Reasons for Delay on TTC
#| 
# Define the order of days
days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

# Convert 'Day' column to a factor with specific levels
clean_transit_data$Day <- factor(clean_transit_data$Day, levels = days_order)

avg_time_delayed_vehicle <- clean_transit_data |> group_by(Day) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = Day, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay by Day of Week", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal()
```

### 2024-01-25 outputs/paper/paper.tex
Cleaned data and started writing paper.

\title{Exploring TTC Delays\thanks{Code and data are available at:
https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé
coffee, Rajan Maghera, Egg McMuffins, Hannah Yu, and Sehar Bajwa for
your love and support - could not have done it without you. No thanks go
to Rohan in particular for not pointing out all the R he wanted us to
learn that was in his textbook\ldots{} sigh. I will try the appendix
cocktails someday, though. :)}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{A Multimodal Analysis}
\author{Luca Carnegie}
\date{January 24, 2024}

@@

A key area of improvement in Canadian sustainability is in increasing
the use of public transportation in big cities like Toronto. This makes
sense since cars account for about 80\% of transportation emissions (Dia
et al. 2019). However, a major drawback preventing the widespread use of
public transportation is the existence of delays within the transit
system. These delays incentivize citizens toward using less sustainable
alternatives like private cars, leading to congestion on highways and
consequent increased emissions. The lack of timely transit services have
a demonstrated impact on traffic. One study by Anderson et al.~estimated
a whopping 47 percent increase in highway delay when transit service
ceased completely in a big city such as Los Angeles (Anderson 2014). It
is clear to see, then, that it is imperative that policy makers
understand and fix these delays to improve sustainability.

The ``TTC'', or Toronto Transit Commission, has been Toronto's transit
system since 1921, serving the city population with subways, streetcars,
buses, along with smaller offerings. The openly available data
surrounding the delays of these various modes of transport offers
opportunity to investigate the root causes of the longer-lasting delays.
Understanding what caused a delay, when it was caused and where it was
caused could give the TTC more insight into how to deliver their
services to the public more efficiently and effectively, setting an
examples of what an effective transit agency could look like.

This paper aims to find relationships and patterns of public transit
delay time across various genres of transportation. First, the data
source and analysis employed are covered. Section Three then critically
examines the data, delving deeper to arrive at various conclusions, as
well as discussing implications and proposing new areas of exploration.

@@

To investigate transit delays in Toronto, data on bus (Commission
2024a), subway (Commission 2024c), and streetcar (Commission 2024b)
delay incidents for the year 2023 was downloaded using the
OpenDataToronto R package (Gelfand 2022). Then, the data was cleaned
using R (R Core Team 2022), tidyverse (Wickham et al. 2019) and it's
associated packages. After combining the datasets together, variables
common between datasets and relevant to the analysis were selected. In
the case of this analysis, the variables included were date (year,
month, day), day of the week, Location of Delay Incident, Cause of
Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as
well as the route/line (simplified to `line') the vehicle was on when
the incident happened. (Perhaps add seasonal analysis??)

\subsection{General Delay Analysis}\label{subsec1-data}

@@

\begin{tabular}{l|l|l|l|r|l|l|r}
\hline
Date & Time & Day & vehicle & Line & Location & Incident & Min Delay\\
\hline
2023-01-01 & 02:30:00 & Sunday & Bus & 91 & WOODBINE AND MORTIMER & Diversion & 81\\
\hline
2023-01-01 & 02:34:00 & Sunday & Bus & 69 & WARDEN STATION & Security & 22\\
\hline
2023-01-01 & 03:06:00 & Sunday & Bus & 35 & JANE STATION & Cleaning & 30\\
\hline
2023-01-01 & 03:14:00 & Sunday & Bus & 900 & KIPLING STATION & Security & 17\\
\hline
2023-01-01 & 03:43:00 & Sunday & Bus & 85 & MEADOWALE LOOP & Security & 1\\
\hline
\end{tabular}

@@

>>>>

\title{Exploring TTC Delays\thanks{Code and data are available at:
https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé
coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and
support - could not have done it without you. No thanks go to Rohan in
particular for not pointing out all the R he wanted us to learn that was
in his textbook\ldots{} sigh. I will try the appendix cocktails someday,
though. :)}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{A Multimodal Analysis}
\author{Luca Carnegie}
\date{January 25, 2024}

@@

A key area of improvement in Canadian sustainability is in increasing
the use of public transportation in big cities like Toronto. This makes
sense, since cars account for about 80\% of transportation emissions
(Dia et al. 2019). A major drawback preventing the widespread use of
public transportation is the existence of delays within transit systems.
These delays push citizens toward using less sustainable alternatives
like private cars, leading to congestion on highways and consequent
increased emissions. The lack of timely transit services have a
demonstrated impact on traffic. One study by Anderson et al.~estimated a
whopping 47 percent increase in highway delay when transit service
ceased completely in a big city such as Los Angeles (Anderson 2014).
Learning from this example, it is imperative that policy makers
understand and fix these delays to improve sustainability in Toronto.

The Toronto Transit Commission (TTC) has been Toronto's transit system
since 1921, serving the city population with a variety of transportation
methods, particularly subways, streetcars, and buses. The openly
available data surrounding the delays of these various modes of
transport offers the opportunity to investigate the root causes of these
delays and prompt discussion as how to rectify them. Understanding the
common causes of a delay, when a delay is typically caused and where it
was caused gives the TTC more insight into how to deliver their services
to the public more efficiently and effectively, setting an improved
example of what an effective transit agency looks like.

This paper aims to find patterns of public transit delay time across
various genres of transportation. First, the data source and analysis
employed are covered in Section Two. Section Three then critically
examines the data, delving deeper to arrive at various conclusions, as
well as discussing implications and proposing new areas of exploration
for this data.

@@

To investigate transit delays in Toronto, data on bus (Commission
2024a), subway (Commission 2024c), and streetcar (Commission 2024b)
delay incidents for the year 2023 was downloaded using the
OpenDataToronto R package (Gelfand 2022). Then, the data was cleaned
using R (R Core Team 2022), tidyverse (Wickham et al. 2019) and it's
associated packages. After combining the datasets together, variables
common between datasets and relevant to the analysis were selected,
leading to a dataset with 93,569 observations. In the case of this
analysis, the variables included were date (year, month, day), day of
the week, Location of Delay Incident, Cause of Delay, the time delay (in
minutes), Vehicle (bus, subway, streetcar), as well as the route/line
(simplified to `line') the vehicle was on when the incident happened.

@@

@@

\begin{table}
\caption{Sample of Cleaned Delay Data}\tabularnewline

\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{l|l|l|l|r|l|l|r}
\hline
Date & Time & Day & Vehicle & Route/Line & Location & Reason & Delay (Minutes)\\
\hline
\cellcolor{gray!6}{2023-01-01} & \cellcolor{gray!6}{02:30:00} & \cellcolor{gray!6}{Sunday} & \cellcolor{gray!6}{Bus} & \cellcolor{gray!6}{91} & \cellcolor{gray!6}{WOODBINE AND MORTIMER} & \cellcolor{gray!6}{Diversion} & \cellcolor{gray!6}{81}\\
\hline
2023-01-01 & 02:34:00 & Sunday & Bus & 69 & WARDEN STATION & Security & 22\\
\hline
\cellcolor{gray!6}{2023-01-01} & \cellcolor{gray!6}{03:06:00} & \cellcolor{gray!6}{Sunday} & \cellcolor{gray!6}{Bus} & \cellcolor{gray!6}{35} & \cellcolor{gray!6}{JANE STATION} & \cellcolor{gray!6}{Cleaning} & \cellcolor{gray!6}{30}\\
\hline
2023-01-01 & 03:14:00 & Sunday & Bus & 900 & KIPLING STATION & Security & 17\\
\hline
\cellcolor{gray!6}{2023-01-01} & \cellcolor{gray!6}{03:43:00} & \cellcolor{gray!6}{Sunday} & \cellcolor{gray!6}{Bus} & \cellcolor{gray!6}{85} & \cellcolor{gray!6}{MEADOWALE LOOP} & \cellcolor{gray!6}{Security} & \cellcolor{gray!6}{1}\\
\hline
\end{tabular}}
\end{table}

\includegraphics{paper_files/figure-pdf/common-reasons-delay-1.pdf}

\begin{verbatim}
Warning: Unknown or uninitialised column: `month`.
\end{verbatim}

\includegraphics{paper_files/figure-pdf/monthly-delay-times-1.pdf}

\includegraphics{paper_files/figure-pdf/mode-longest-delays-1.pdf}

\includegraphics{paper_files/figure-pdf/most-incidents-vehicle-1.pdf}

\includegraphics{paper_files/figure-pdf/transport-delay-time-1.pdf}

\includegraphics{paper_files/figure-pdf/transport-delay-number-1.pdf}
